---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    fig_caption: yes
    keep_md: yes
    
---


#### Loading and preprocessing the data
```{r}
data <-  read.csv("activity.csv", header = TRUE, sep = ",", na.strings = "NA")
```

 Converting the "date" variable to a Date class and "interval" variable to factor
```{r}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
data$interval <- factor(data$interval)
```

#### Remove NA Values
```{r}
NA_index <- is.na(as.character(data$steps))
data_no_NA <- data[!NA_index,]
```

#### Creating a data frame with the steps taken for each day
```{r}
steps_per_day <- aggregate(steps ~ date, data = data_no_NA, sum)
```
#### Adding column names to the created data frame
```{r}
colnames(steps_per_day) <- c("date", "steps")
```
# What is mean total number of steps taken per day?
```{r, echo=FALSE}
hist(as.numeric(steps_per_day$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Histogram of the total number of steps taken per day")
```

## Mean total number of steps taken per day
```{r}
mean(steps_per_day$steps)
```
## Median total number of steps taken per day
```{r}
median(steps_per_day$steps)
```

# What is the average daily activity pattern?
#### Calculating the average
```{r}
steps_per_interval <- aggregate(data_no_NA$steps, by=list(interval=data_no_NA$interval), FUN=mean)
```

#### Adding columns names
```{r}
colnames(steps_per_interval) <- c("interval", "average_steps")
```

#### ploting the average daily activity pattern 
```{r}
plot(as.integer(levels(steps_per_interval$interval)), steps_per_interval$average_steps, type="l",
     xlab = "Interval", ylab = "Average Number of Steps", main = "Average Daily Activity Pattern",  col ="blue")
```

#### The 5-minute interval that contains the maximum number of steps
```{r}
interval_max_steps<-steps_per_interval[which.max(steps_per_interval$average_steps),]$interval
interval_max_steps
```

# Imputing missing values
#### Number of missing values
```{r}
missing_values<-sum(is.na(data))
missing_values
```
#### Fill up missing values with mean
#### Finding the indices of missing values (NAs)
```{r}
NA_index <- which(is.na(as.character(data$steps)))
complete_data <- data
```
#### Inputing missing values using the mean for that 5-minute interval
```{r}
complete_data[NA_index, ]$steps<-unlist(lapply(NA_index, FUN=function(NA_index){
                steps_per_interval[data[NA_index,]$interval==steps_per_interval$interval,]$average_steps
                }))
```
## Displaying the new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
summary(complete_data)
str(complete_data)
```
#### Creating a data frame with the steps taken for each day
```{r}
steps_each_day_complete <- aggregate(steps ~ date, data = complete_data, sum)
```
#### Adding column names to the created data frame
```{r}
colnames(steps_each_day_complete) <- c("date", "steps")
```
## Making the histogram and report mean and median
```{r, echo=FALSE}
hist(as.numeric(steps_each_day_complete$steps), breaks = 20, col = "red", xlab = "Number of Steps", main= "Histogram of the total number of steps taken each day")
```
#### Mean
```{r}
mean(steps_each_day_complete$steps)
```
##### Median
```{r}
median(steps_each_day_complete$steps)
```
## Are there differences in activity patterns between weekdays and weekends?
 
#### Creating a factor "day "to store the day of the week
```{r}
complete_data$day <- as.factor(weekdays(complete_data$date))
```
#### Creating a logical variable "is_weekday" where weekday=TRUE, weekend = FALE
```{r}
complete_data$is_weekday <- ifelse(!(complete_data$day %in% c("Saturday","Sunday")), TRUE, FALSE) 
```

#### Calculating average number of steps for weekdays
```{r}
weekdays_data <- complete_data[complete_data$is_weekday,]
steps_per_interval_weekdays <- aggregate(weekdays_data$steps, by=list(interval=weekdays_data$interval), FUN=mean)
```

#### Calculating average number of steps for weekends
```{r}
weekends_data <- complete_data[!complete_data$is_weekday,]
steps_per_interval_weekends <- aggregate(weekends_data$steps, by=list(interval=weekends_data$interval), FUN=mean)
```
#### Adding columns names
```{r}
colnames(steps_per_interval_weekdays) <- c("interval", "average_steps")
colnames(steps_per_interval_weekends) <- c("interval", "average_steps")
```
#### Adding a column to indicate the day
```{r}
steps_per_interval_weekdays$day <- "Weekday"
steps_per_interval_weekends$day <- "Weekend"
```
#### Merging the two together
```{r}
week_data <- rbind(steps_per_interval_weekends, steps_per_interval_weekdays)
```
#### Converting the day variable to a factor
```{r}
week_data$day <- as.factor(week_data$day)
```
## Making the plot
```{r}
library(lattice)
xyplot(average_steps ~  interval | day, data = week_data, layout = c(1,2), type ="l", ylab="Number of Steps")
```